The MRChem MRA code for molecular electronic calculations. Performance and scaling properties

### 2022-05-11 main.tex
JCP template

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{braket}
\usepackage{natbib}
\usepackage{graphicx}

\begin{document}

\input{newcommands.tex}

\title{The MRChem MRA code for molecular electronic calculations. Performance and scaling properties}

\author{Peter Wind, Magnar Bjørve, Anders Brakestad, Gabriel Gerez, Stig Rune Jensen, Roberto Di Remigio and Luca Frediani}

%\affilition{Department of Chemistry, UiT - The Arctic University of Norway, N-9037 Tromsø, Norway}

@@

\begin{abstract}
MRChem is a Multiresolution Analysis (MRA) code for molecular electronic calculations, based on a multiwavelets adaptive basis representation. A description of the implementation strategy is provided. Benchmark results on systems comprising more than thousand orbitals at the Hartree Fock (HF) level are analyzed, with an emphasis on scaling properties. 
We show that some terms which formally scale as $~0(N^2)$, in effect have a better scaling because of the implicit screening introduced by the inherent adaptivity of the method. Comparison with traditional GTO based software, show that MRChem can be competitive wrt computation time.

%test

@@

Gaussian basis sets (GTO and more generally Linear Combination of Atomic Orbitals, LCAO) are well established as a standard for ab initio molecular calculations. 
Even very small basis sets (a few functions per orbital) can give reasonable results for describing molecular properties.

% Husker ikke om jeg skrev dette selv eller om det er tatt fra en artikkel!: In a localized orbital perspective, atom centered basis set are very effective to describe orbitals belonging to that atom, but are basis functions centered on other atoms will be much less useful for describing the local orbitals. For systems with many atoms, one ends up with describing orbitals with a large proportion of very small, coefficients. Those small parts are still necessary among other in order to ensure orthogonality among orbitals. The natural solution is then to describe different orbital or functions in different sets of basis functions. There are many problems associated with this approach, the main one is that each mathematical operation involving more than one function, must relate different basis sets and a new basis has to be defined for every new resulting function. It is a challenge to do this in a balanced and efficient way.

In a MRA framework, the basis can adapt according to the function described (for an in-depth review of the MRA method in the field of Quantum Chemistry, see \citep[]{bischoff2019}). The available basis is in principle infinite, and the basis actually used is dynamically adapted to the required precision. However reasonable results can require the real-space basis to comprise millions of elements for each function. In this sense, the method starts with a big handicap, compared to Gaussian basis sets.

The challenge is not only the large amount of operations needed to perform calculations on large systems, but also the large memory footprint. Using a computer cluster, if not all the data can be stored locally, the data must be stored on separate compute nodes. On modern computers access to data is generally more time consuming than the mathematical operations itself, and if the data is not available locally, this is even more important. The implementation must be able to efficiently copy large amounts of data between compute nodes and the algorithm must be devised so as to use or reuse the data already locally available when possible.

@@

%For larger systems, or high precision, traditional basis set methods require large basis sets, as the same basis is used for all orbitals. Methods exists to "screen" the basis in order to use only a limited part of the entire basis, however the screening will only take effect if the corresponding part of the density matrix can be neglected. Higher accuracy requires larger basis sets, {\sl and} that a larger part of the space must be considered. For those systems the advantages of the Gaussian basis is not obvious.

In this article, we will present the main implementation features of a MRA code, MRChem, which at present is capable of describing systems with thousands of orbitals at the Hartree Fock or Density Functional Theory (DFT) level.

The large memory requirement is addressed by storing the data in a distributed "memory bank", where orbitals can be sent and retrieved independently by any CPU. The total size of the memory bank is then only limited by total size available on the entire computer.

The large computational cost is addressed by parallelization either at the orbital level or through a partition of the real space. This  dual parallelization approach, allows to minimize the relative cost of communication and the overall computational costs.

@@

Using localized orbitals the adaptiveness of the multiwavelets description will reduce significantly the computational effort required to compute the terms that involves remote orbitals. This opens the way for a linearly scaling method where the linearity arises naturally from the methodology.

Benchmark calculations at the Hartree Fock level show that MRChem is able to describe systems with thousands of electrons, and exhibits near linear scaling properties. Computation times are competitive with state of the art traditional software.

\section{The multiwavelet method for Quantum Chemistry}

Only a superficial overview of the method is given here, as more detailed descriptions can be found elsewhere REF.

The Kohn Sham or HF equations can be written

\begin{equation}
  (-\frac{1}{2}\nabla^2 + V) \psi_i = \sum_{j_{occ}} F_{ij} \psi_j
\end{equation}

\begin{equation}
  F_{ij} = \langle \psi_i |{-\frac{1}{2}\nabla^2+V}| {\psi_j}\rangle
\end{equation}

In the HF case $V=V_{nuc}+J-K$, and in the Kohn Sham equations $V=V_n+J+V_{xc}$, where $V_{nuc}$, $J$, $K$ , $V_{xc}$ are the nuclear, Coulomb, Exchange and exchange-correlation potentials respectively. $F$ is the Fock matrix and $-\frac{1}{2}\nabla^2 = T$ is the kinetic operator.

Compared to more standard LCAO methods, there are a few limitations introduced by the wavelet approach:

Since the  MW functions are not continuous, derivation is problematic. The direct application of the kinetic operator, will not give an accurate result (but good approximations can be found [ref]).

In order to avoid direct use of the Kinetic operator one rewrite the equation as,

\begin{equation}
\psi_i = -2 G_{i}(V\psi_i - \sum_{j_{occ} \ne i} F_{ij} \psi_j)
\end{equation}

where the Helmholtz operator $G_{i}$ is defined by
\begin{equation}
G_{i}(-\nabla^2 - 2 F_{ii} )=1
\end{equation}

Since the number of mw basis is not a priori limited, there is no natural way of defining a set of "virtual orbitals". We cannot define the Fock matrix in the entire basis and even less diagonalize it.

\subsection{Basic scf operations} %put complex conjugates!

We will here show the main steps involved in a SCF iteration in a mw framework.

First the matrix elements of the Fock operator are evaluated explicitly. This involves the application of the kinetic operator ($T$), nuclear  potential ($V_{nuc}$), Coulomb operator ($J$), and  the exchange potential ($K$) and/or the "exchange-correlation" potential from the density functional ($V_{xc}$) on each occupied orbital.

\subsubsection{$T$}
As mentioned the application of the kinetic operator on a mw function can be problematic. However first order derivatives are less problematic than second order derivatives. Further, since the matrix element involves an integration, the resulting numbers are in practice accurate enough. 
In our implementation the matrix elements are computed by first evaluating the first order derivative and the overlap between them:
\begin{equation}
  T_{ij} = -\frac{1}{2} \langle \nabla \psi_i |\nabla {\phi_j}\rangle
\end{equation}

@@

The application of the nuclear potential (and possibly additional external fields) can be performed in two ways. Either the sum over nuclei is performed as a sum of matrix elements for the individual nuclei, or the sum over the potentials is performed first, and the matrix elements evaluated at the end. In our present implementation, the first form is used.
\begin{equation}
  V_{nuc} \phi_j = (\sum_{k = nuclei} V_k) \phi_j
\end{equation}
The total potential for all nuclei ($\sum_{k} V_k$) can be very large in terms of memory. It describes the potential over the entire space, whereas only the parts that overlap with the occupied orbitals are actually used. In the future we plan to adapt the accuracy according to the electronic density.

@@

\subsubsection{$J$}

We compute the total density before the total Coulomb potential ($V_C$) is constructed and applied to the orbitals:

\begin{equation}
  V_{C}(r) = \int \frac{\sum_k \phi^*_k(r')\phi_k(r')}{|r-r'|} \ud r'
\end{equation}

\subsubsection{$K$}

The exchange terms are probably the most computationally demanding parts, because they require the application of the Coulomb operator for every orbital pair:

\begin{equation}
  K_{j} = \sum_k  K_{Ex}^k \phi_j 
\end{equation}
with 
\begin{equation}
  K_{Ex}^k(r) \phi_j = \phi_k(r) \int \frac{\phi^*_k(r')\phi_j(r')}{|r-r'|} \ud r'
\end{equation}

Section \ref{Xscreen} provides a more detailed description of how these terms can be evaluated efficiently.

\subsubsection{$V_{xc}$}
In our implementation, the value of the density is evaluated at each point in space, and the dft potential is then evaluated using the external library XCfun[REF], and transformed back into a mw potential.

@@

There is one Helmholtz operator associated to each orbital energy, however each Helmholtz operator is applied to a single orbital.

\subsubsection{Basic mw operations required}

The elementary mw operations required for all those steps are
\begin{itemize}
\item Construction of a mw function given an analytic or sampled function.
\item Evaluation of the value of a mw function at a given point in space.
\item Multiplication by a number.
\item Sum of two mw functions.
\item Product of two mw functions.
\item Overlap between two mw functions.
\item Application of a potential on a mw function (multiplication).
\item Evaluation (approximative) of the first order derivative of a mw function.
\item Application of Coulomb operator on a mw function to get a potential.
\item Construction and application of the Helmholtz operator.
\end{itemize}
All these operations are provided by the MRCPP library.

In addition the SCF requires schemes, for orthonormalization, localization and for making the guess of the next orbitals given the preceding iteration results (REF, KAIN).

\subsection{MRChem software, MRCPP and Vampyr}
%section about availability and things the code can do for the user
The MRChem code relies on the MRCPP library for performing the basic mathematical operations in a multiwavelet basis.

git, properties, Vampyr ...

@@

MW calculations can be computationally demanding. Both the total amount of elementary operations and the large amount of memory required and the capacity of the network are important bottlenecks.

In practice a computer cluster is required for calculations on large systems.
For example the representation of one orbital is described typically by of the order of 10-500MB of data (see section \ref{sizes}). Usually several sets of orbitals are required in the scf process (previous iteration, right hand size of equations etc.). If 1000 orbitals are described, this will often require more memory than is locally available on a compute node of a cluster. Therefore it is required to store the data in a distributed way. At the same time, the equations require to access all pairs of orbitals, i.e. all the data must at some point be accessible locally.

@@

Although localized and canonical orbitals represent the same wave function, the numerical operations required to solve the equations will have different scaling properties in a canonical or localized representation.
Because of the inherent adaptiveness of the mw approach, multiplying two orbitals, or applying an operator onto an orbital can be done more efficiently if the orbitala are localized in space. Even if the orbitals have non-negligible values in a large portion of space, the adaptiveness properties will make sure that the calculations in the parts where the orbitals are small are done at a coarser level of accuracy, and therefore faster than in the regions with larger values.

\subsection{Adaptivity}
%explanation of implicit screening mechanisms within MRA? 
% i.e. MRCPP tree calculation using norm threshold

@@

The orbitals need to be stored on the computer. This can be done in several ways. The KS or HF equations are written in terms of orbitals, and therefore it is natural to implement the code in a way that follows the mathematical expressions, i.e. each orbitals is stored individually. Another way of partitioning the problem, is to consider one part of real space at a time, and store together the contributions from all the orbitals at that position. Both approaches have advantages and disadvantages.

However we will show here why it is advantageous, for some terms at least, to shift to a real-space partitioning approach.

(This is similar to the history for gaussian basis set methods, which started using Molecular Orbitals, but for large systems could get more efficient implementations by expressing the core terms using an "Atomic Orbitals" representation (Helgaker, T.; Jørgensen, P.; Olsen, J. Molecular Electronic-Structure Theory; John Wiley \& Sons, Ltd, 2014).)

In the computation of the Fock matrix elements, one has to evaluate terms of the form

\begin{equation}
  {\hat O}_{ij} = \braket{\phi_i | {\hat O} |\phi_j}  = \braket{\phi_i | {\hat O} \phi_j} 
\end{equation}

where $\phi_i$ and $\phi_j$ are occupied orbital, and ${\hat O}$ is an operator, for example the nuclear potential operator. The application of the operator is a $~0(N)$ operation, and the calculations of the entire matrix scales formally as $~0(N^2)$. We can express explicitly the orbitals in the multiwavelet basis:

\begin{equation}
  \phi_i = \sum_{k,n} c_{ikn} p_{kn} 
\end{equation}
\begin{equation}
  {\hat O} \phi_j = \sum_{k,n} d_{jk_n} p_{kn} 
\end{equation}

where $p_{kn}$ is a member of the used mw basis, i.e. a three-dimensional polynomial (index $k$) for a multiwavelet node (index $n$) at specific position in space, and $c_{jkn}$ are the coefficients that define the function $\phi_j$ in that basis.

Since the basis is adaptive, different functions will use different sets of nodes. 
In order to compute the overlap of two functions $\phi_i$ and $\phi_j$, they are first adaptively projected onto a common basis and multiplied.

\begin{equation}
  \braket{\phi_i|{\hat O} \phi_j} = \sum_{k_n} c_{ikn}d_{jkn}
\end{equation}

@@

This operation is equivalent to the multiplication of the $c$ and $d$ matrices. However for large systems, it is not possible to hold the coefficients for all the indices i,j,k,n in local memory.
The "MO" approach is then to sum over all $k,n$ for for each pair ij independently. The problem is that each coefficient $c_{jkn}$ is used only once at a time. On modern based computers memory access are much more time demanding (factor of the order of 100 for most CPU based systems) than the mathematical operations itself.  A much faster implementation is to consider a specific node n and perform the product for all i, j and k values within this node.

\begin{equation}
  \braket{\phi_i | {\hat O} |\phi_j}_n = \sum_{k} d_{jkn} p_{kn} \forall i,j
\end{equation}

@@

@@

MRChem is parallelized using Message Passing Interface (MPI) and OpenMP.

@@

The low level (operations on one or two orbitals or mw trees, and on nodes) are included  in the MRCPP library. The MRCPP will perform tree operations as a set of operations on nodes. Those operations will then be performed in parallel by OpenMP threads. The thread parallelity is transparent for the library user; only the number of threads needs to be set.

@@

In the multiwavelet approach, orbitals are functions represented in a tree structure. The root nodes cover the entire space of the system, and each subdivision defines branches. In order to facilitate the transfer of those functions, the tree structure and the node data are stored in piecewise contiguous memory. This allows the data to be sent efficiently among MPI processes: only the position in memory of the start and end are required, and some metadata information. A tree can then be transfered without intermediate buffering and with little overhead. 
The MRCPP library contains functionalities to easily and efficiently transfer mw trees (or orbitals) between MPI processes.

@@

\subsubsection{Task manager} 
In a localized orbital approach, many of the elements of the matrix $\braket{\phi_i|{\hat O} \phi_j}$ will be small. Since the method is inherently adaptive using the MRCPP library, the small elements will be faster to compute. In order to fully exploit this feature, a dynamical distribution of tasks is required.  
In practice the matrix $\braket{\phi_i|{\hat O} \phi_j}$ is partitioned into blocks. Each block represent one task to be performed by a "worker". In a local representation, each node is simply one task. When a "worker" is done with one task it can ask for a new task from the "task manager" until all the tasks are finished.

One single MPI process is reserved for managing the task queue. The only purpose of this process is to keep account of the tasks that have been assigned and the tasks left, and distribute unsolved tasks to the workers. It does only send and receive small amounts of data at a time (few bytes) and will therefore in practice not be saturated with requests.

@@

\begin{equation}
\label{equ:ex}
  K_{Ex}^k(r) \phi_j = \phi_k(r) \int \frac{\phi^*_k(r')\phi_j(r')}{|r-r'|} \ud r'
\end{equation}

@@

The evaluation of equation (\ref{equ:ex}) can be divided into 3 operations:
\begin{itemize}
\item Product of $\phi^*_k(r')$ and $\phi_j(r')$
\item Application of Poisson operator, $\int \frac{(\phi^*_k(r')\phi_j(r'))}{|r-r'|} \ud r'$
\item Product with $\phi_k(r)$, $\phi_k(r) (\int \frac{\phi^*_k(r')\phi_j(r')}{|r-r'|} \ud r')$
\end{itemize}

@@

The scheme for orthogonalization of orbitals uses the diagonalization of a hermitian matrix and subsequent matrix multiplications which both scales as $N_{occ}^3)$

>>>>

\documentclass[%
 aip,
% jmp,
% bmf,
% sd,
% rsi,
 amsmath,amssymb,
%preprint,%
 reprint,%
%author-year,%
%author-numerical,%
% Conference Proceedings
]{revtex4-1}

\usepackage{graphicx}% Include figure files
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{bm}% bold math
%\usepackage[mathlines]{lineno}% Enable numbering of text and display math
%\linenumbers\relax % Commence numbering lines

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{mathptmx}
\usepackage{etoolbox}

%PW additional packages
\usepackage{braket}
\usepackage{todonotes}

%% Apr 2021: AIP requests that the corresponding 
%% email to be moved after the affiliations
\makeatletter
\def\@email#1#2{%
 \endgroup
 \patchcmd{\titleblock@produce}
  {\frontmatter@RRAPformat}
  {\frontmatter@RRAPformat{\produce@RRAP{*#1\href{mailto:#2}{#2}}}\frontmatter@RRAPformat}
  {}{}
}%
\makeatother
\begin{document}

\preprint{AIP/123-QED}

\title[MRChem]{The MRChem MRA code for molecular electronic calculations. Performance and scaling properties}
\author{Peter Wind}
\affiliation{Department of Chemistry, UiT - The Arctic University of Norway, N-9037 Tromsø, Norway}
\author{Magnar Bjørve}
\affiliation{Department of Chemistry, UiT - The Arctic University of Norway, N-9037 Tromsø, Norway}
\author{Anders Brakestad}
\affiliation{Department of Chemistry, UiT - The Arctic University of Norway, N-9037 Tromsø, Norway}
\author{Gabriel Gerez}
\affiliation{Department of Chemistry, UiT - The Arctic University of Norway, N-9037 Tromsø, Norway}
\author{Stig Rune Jensen}
\affiliation{Department of Chemistry, UiT - The Arctic University of Norway, N-9037 Tromsø, Norway}
\author{Roberto Di Remigio Eikås}
\affiliation{ENCCS Department of Information Technology, Uppsala University, Uppsala, Sweden}
%To check Roberto affiliation
\author{Luca Frediani}
\affiliation{Department of Chemistry, UiT - The Arctic University of Norway, N-9037 Tromsø, Norway}

\date{\today}

\begin{abstract}
MRChem is a MRA code for molecular electronic structure calculations, based on a multiwavelets adaptive basis representation. A description of the implementation strategy is provided. Benchmark calculations on systems comprising more than thousand orbitals at the HF level are analyzed, with an emphasis on scaling properties. 
We show that some terms which formally scale as $~0(N^2)$, in effect have a better scaling because of the implicit screening introduced by the inherent adaptivity of the method. Comparison with traditional GTO based software, show that MRChem can be competitive with respect to computation time.

@@

@@

GTO and more generally MRA are well established as a standard for \emph{ab-initio} molecular electronic structure calculations. As their shape is closely related to the electronic structure of atoms, even very small basis sets (a few functions per orbital) can give reasonable results for describing molecular properties. However for extended systems the description of each orbital still requires the contributions from the entire basis in order to ensure orthogonality. Even when using localized orbitals, a large proportion of the coefficients will be very small for those systems.

In a MRA framework, the basis can adapt according to the function described (for an in-depth review of the MRA method in the field of Quantum Chemistry, see \citep[]{bischoff2019}). The available basis is in principle infinite and complete, and the basis actually used is dynamically adapted to the local shape of the function and the required precision. However reasonable results can require the real-space basis to comprise millions of elementary functions for each orbital. In this sense, the method starts with a big handicap, compared to Gaussian basis sets.

The challenge is not only the large amount of operations needed to perform any mathematical transformation in this approach, but also the large memory footprint: beyond a certain system size, not all data can be stored in local (i.e. fast access) memory. On modern computers, data access is generally more time consuming than the mathematical operations itself, especially if the data is not available locally. The implementation must be able to efficiently copy large amounts of data between compute nodes and the algorithm must be devised so as to use or reuse the data already locally available when possible.

@@

In this article, we will present the main implementation features of a MRA code, MRChem, which is capable of describing such large systems at the Hartree Fock or Density Functional Theory (DFT) level.

The large memory requirement is addressed by storing the data in a distributed "memory bank", where orbitals can be sent and retrieved independently by any CPU. The total size of the memory bank is then only limited by overall memory available on the entire computer cluster.

The large computational cost is addressed by parallelization either at the orbital level or through real space partitioning. This dual strategy allows to minimize the relative cost of communication and the overall computational costs. Further the most computational intensive parts are expressed as matrix multiplications, which allows for optimal efficiency.

@@

Using localized orbitals the adaptivity of the multiwavelets description will significantly reduce the computational effort required to compute the terms involving remote orbitals. This opens the way for a linearly scaling method where the linearity arises naturally from the methodology, rather than being obtained by specific additional methods such as multipole interactions or purification of the density matrix \cite{trygve}.

Benchmark calculations at the Hartree Fock level show that MRChem is able to describe systems with thousands of electrons, exhibits near-linear scaling properties and can also be competitive with state-of-the-art electronic structure software based on LCAO methods.

\section{Solving the Hartree-Fock and Kohn-Sham equations with Multiwavelets}

We consider the SCF equations common to the HF method and the KS equations of DFT:

\begin{equation}\label{eq:differential-scf}
  \left( -\frac{1}{2}\nabla^2 + V \right) \psi_i = \sum_{j_{occ}} F_{ij} \psi_j
\end{equation}

where the Fock matrix elements $F_{ij}$ are obtained as
\begin{equation}\label{eq:fock-matrix-definition}
  F_{ij} = \braket{ {\psi_i}|{-\frac{1}{2}\nabla^2+V}|{\psi_j}}
\end{equation}

In the above equations, $T = -\frac{1}{2}\nabla^2$ is the kinetic energy operator, and the potential $V$ contains the nuclear-electron attraction $V_{nuc}$, the Coulomb repulsion $J$, the exact exchange $K$ and, in case of DFT, the exchange and correlation potential $V_{XC}$.

To solve the SCF equations within a MW framework, it is necessary to rewrite the differential equation \ref{eq:differential-scf} as an integral equation:
\begin{equation}\label{eq:integral-scf}
\psi_i = -2 H^{\mu_i}\left[V\psi_i - \sum_{j_{occ} \ne i} F_{ij} \psi_j\right]
\end{equation}
This is obtained by recalling that the shifted kinetic energy operator $T-\epsilon$ has a closed-form Green's function:
\begin{equation}\label{eq:helmholtz-green}
    (T-\epsilon)^{-1} = 2 H^{\mu}(r) \quad H^{\mu}(r) = \frac{e^{-\mu r}}{r} \quad \mu = \sqrt{-2\epsilon} 
\end{equation}
%Working in the integral form is both a necessity and an advantage. %\todo[inline]{PW: I do not see why it is an advantage; it would be much easier if we could use a basis with an exact second order derivative}

A multiwavelet representation does not have a predefined basis, and as such it is not possible to express functions and operators as vectors and matrices of a predefined size and the virtual space is to be regarded as infinite, although each function has a finite, truncated representation defined by the chosen precision threshold. It is therefore necessary to have working equations which allow the direct optimization of orbitals.
At the same time one deals only with occupied orbitals and the Fock matrix in Equations \ref{eq:differential-scf} and \ref{eq:integral-scf} is therefore limited in size.

Differential operators such as the Laplacian, also pose a fundamental problem for function representations which are not continuous, and a naive implementation leads to numerical instabilities. Robust approximation are nowadays available\cite{der-paper}, but for repeated iterations avoiding the use of the Laplacian operator is still an advantage.

%\todo[inline]{PW: The "preconditioner" is only useful if it accelerate the convergence. The ideal (or at least exact to second order) preconditioner is the inverse of the Hessian. In traditional methods one use good approximations for the inverse Hessian (diagonal or so), while the "preconditioner" we use is only a matrix that we put in front because it is the only way we can practically solve the equation, but is not optimized for being a good approximation of the inverse Hessian. I have therefore rewritten the text below}

Additionally, it can be shown that iterating on the integral equation corresponds to a preconditioned steepest descent. Practical applications show that this approach is comparable in efficiency (as measured in the rate of convergence of the SCF iterations) to more traditional methods [REF?].

\subsection{The basic operations in the SCF iterations} %put complex conjugates!

For a detailed description of the SCF optimization using Multiwavelets, the reader is referred to a recently published article which describes the algorithm in detail [REF].

We will here list the main operations, with emphasis on those aspects which are relevant for an efficient parallelization.

The starting point is an initial set of orthonormal orbitals which define a Slater determinant for the given problem. A density is computed and the corresponding operators ($J$, $K$ and $V_{XC}$) are built. The Fock matrix (Eq.~\ref{eq:fock-matrix-definition} is computed. This requires the application of the kinetic operator ($T$), nuclear  potential ($V_{nuc}$), Coulomb operator ($J$), and  the exchange potential ($K$) and/or the "exchange-correlation" potential from the density functional ($V_{xc}$) on each occupied orbital. The argument for the application of the integral operator is assembled (the square bracket in Eq.~\ref{eq:integral-scf}). Then the integral operator is applied and the cycle is iterated after a normalization step.

The elementary operations required for all those operations are
\begin{itemize}
\item Construction of a MW function given an analytic or sampled function.
\item Evaluation of the value of MW functions at a given point in space.
\item Multiplication of a function by a number.
\item Sum of two MW functions.
\item Product of two MW functions.
\item Computation of the overlap integral between two MW functions.
\item Application of a potential on a MW function (multiplication).
\item Evaluation of the first order derivative of a MW function.
\item Application of Coulomb operator on a MW function to obtain an electrostatic potential.
\item Construction and application of the Helmholtz operator.
\end{itemize}

All these operations are provided by the MRCPP library\cite{mrcpp} and must be composed to achieve a SCF implementation, and they must be integrated in a parallel framework to achieve good scaling and performance.

In addition, the SCF cycle requires schemes for orbital orthonormalization, rotation, and localization, Fock matrix diagonalization, history storage and utilization to enable KAIN acceleration.~\cite{kain}

\subsubsection{$T$}
The application of the kinetic operator on a MW function is formally not well defined. However, it is possible to compute first derivatives provided repeated applications are avoided. The matrix elements of the kinetic energy operator can therefore be computed as:
\begin{equation}
  T_{ij} = -\frac{1}{2} \braket{\nabla \psi_i | \nabla \psi_j}
\end{equation}
Computing the matrix elements require therefore only the gradient followed by the inner product. This last step ensures that the resulting number is in practice accurate enough.\cite{derivative-paper}

@@

\todo[inline]{I'm not quite sure I have understood this explanation...}
\todo[inline]{PW: it is more explicit now, better?}

The application of the nuclear potential (and possibly any additional external field) can be performed in two ways. Either the sum over nuclei is performed as a sum of matrix elements for the individual nuclei:
\begin{equation}
  V_{nuc} \phi_j = \sum_{k = nuclei} (V_k \phi_j)
\end{equation}

or the sum over the potentials is performed first, and the matrix elements evaluated at the end:
\begin{equation}
  V_{nuc} \phi_j = (\sum_{k = nuclei} V_k) \phi_j
\end{equation}
In our present implementation, the second form is used.

The total potential for all nuclei ($\sum_{k} V_k$) can be very large in terms of memory. It describes the potential over the entire space, whereas only the parts that overlap with the occupied orbitals are actually used. In the future we plan to adapt the accuracy according to the electronic density.

@@

\subsubsection{The Coulomb potential $J$}

The coulomb interaction is computed in two steps. First the total density is obtained as a sum of orbital contributions. Thereafter the Coulomb potential is computed by applying the Poisson kernel.
\begin{eqnarray}
  \rho(r) &= \sum_k \phi^*_k(r)\phi_k(r)\\
  J(r) &= \int \frac{\rho(r')}{|r-r'|} \, dr'
\end{eqnarray}

\subsubsection{The exchange potential $K$}

The exchange potential is the most computationally demanding part, because it cannot be written in closed form. A Coulomb potential must be constructed and applied for each orbital pair:

\begin{equation}
  K \phi_j = \sum_k  K_k \phi_j 
\end{equation}
with 
\begin{equation}
  K_k \phi_j = \phi_k(r) \int \frac{\phi^*_k(r')\phi_j(r')}{|r-r'|} \, dr'
\end{equation}

Section \ref{Xscreen} provides a more detailed description of how these terms can be evaluated efficiently in a MRA framework.

\subsubsection{$V_{xc}$}
\todo[inline]{PW: "each point in space" is not very correct. Should be "at the same resolution as the used basis". Reformulate? }
The density is evaluated at each point in space, and the exchange and correlation potential $V_{XC}$ is then evaluated using the external library XCfun[REF], and transformed back into a MW potential.

@@

The Helmholtz operator defined in Eq.~\ref{eq:helmholtz-green} and employed in Eq.~\ref{eq:integral-scf}, depends on the orbital energy. As such one needs to construct as many operators as there are orbitals.

%\subsection{MRChem software, MRCPP and Vampyr}
%section about availability and things the code can do for the user
%\todo[inline]{Do we nee this at all?}
%\todo[inline]{PW: probably not if there is a reference to another paper where this is shown}

@@

MW calculations can be computationally demanding. Both the total amount of elementary operations, the large amount of memory required and the capacity of the network are important bottlenecks.

In practice a supercomputer is required for calculations on large systems.
For example the representation of one orbital is described typically by of the order of 10-500MB of data (see section \ref{sizes}). Usually several sets of orbitals are required in the scf process (previous iterations, right hand size of equations etc.). If 1000 orbitals are described, this will often require more memory than is locally available on a single compute node in a cluster. Therefore it is required to store the data in a distributed way. At the same time, the equations require to access all pairs of orbitals, i.e. all the data must at some point be accessible locally.

@@

Although localized and canonical orbitals represent the same N-electrons wave function, the numerical operations required to solve the equations will have different scaling properties in a canonical or localized representation.
Because of the inherent adaptiveness of the MW approach, multiplying two orbitals, or applying an operator onto an orbital can be done more efficiently if the orbitals are localized in space. Even if the orbitals have non-negligible values in a large portion of space, the adaptiveness properties will make sure that the calculations in the parts where the orbitals are small are done at a coarser level of accuracy, and therefore faster than in the regions with larger values.

@@

The orbitals need to be stored on the computer. This can be done in several ways. The KS or HF equations are written in terms of orbitals, and therefore it is natural to implement the code in a way that follows the mathematical expressions, i.e. each orbital is stored individually. Another way of partitioning the problem, is to consider one part of real space at a time, and store together the contributions from all the orbitals at that position. Both approaches have advantages and disadvantages.

We will show here why it is advantageous, for some terms at least, to shift to a real-space partitioning approach.

(This is similar to the history for gaussian basis set methods, which started using Molecular Orbitals, but for large systems could get more efficient implementations by expressing the core terms using an "Atomic Orbitals" representation \cite{trygve})

In the computation of the Fock matrix elements, one has to evaluate terms of the form 
%note: the equation is not interpreted right by overleaf if physics package is used
\begin{equation}
  {\hat O}_{ij} = \braket{\phi_{i} | {\hat O} |\phi_j}  = \braket{\phi_i | {\hat O} \phi_j} 
\end{equation}

where $\phi_i$ and $\phi_j$ are occupied orbital, and ${\hat O}$ is an operator, for example the nuclear potential operator. If the application of the operator is considered as a $~0(N)$
%not if the operator grows with N. Not that simple!
operation, the calculations of the entire matrix runs over all the orbital pairs and therefore scales formally as $~0(N^2)$. We can express explicitly the orbitals in the multiwavelet basis:

\begin{equation}
  \phi_i = \sum_{k,n} c_{ikn} p_{kn} 
\end{equation}
\begin{equation}
  {\hat O} \phi_j = \sum_{k,n} d_{jkn} p_{kn} 
\end{equation}

where $p_{kn}$ is a member of the used MW basis, i.e. a three-dimensional polynomial (index $k$) for a multiwavelet node (index $n$) at specific position in space, and $c_{ikn}$ are the coefficients that define the function $\phi_i$ in that basis.

Since the basis is adaptive, different functions will use different sets of nodes. 
In order to compute the overlap of two functions $\phi_i$ and ${\hat O} \phi_j$, they are first adaptively projected onto a common basis and multiplied:

\begin{equation}
  \braket{\phi_i|{\hat O} \phi_j} = \sum_{k,n} c_{ikn}d_{jkn}
\end{equation}

@@

This operation is equivalent to the multiplication of the $c$ and $d$ matrices. However for large systems, it is not possible to hold the coefficients for all the indices i,j,k,n in local memory.
The "MO" approach is then to sum over all $k,n$ for each pair ij independently. The problem is that each coefficient $c_{ikn}$ is used only once at a time. On modern based computers memory access are much more time demanding (factor of the order of 100 for most CPU based systems) than the mathematical operations itself.  A much faster implementation is to consider a specific node n and perform the product for all i, j and k values within this node.

\begin{equation}
\label{equ:sum}
  \braket{\phi_i | {\hat O} |\phi_j}_n = \sum_{k} c_{ikn} d_{jkn} \forall i,j
\end{equation}

@@

\subsection{Adaptivity}

In the MRA framework, the required precision can limit the number of elementary operations at several levels.

At the lowest level, most operations on a function in a multiwavelet representation will successively refine the local basis (i.e. grid) of the resulting function, until the required precision is reached. That means that the computational effort required will vary according to the position and shape of the functions or operators. For example the product of two functions which are localized in different part of space, will normally be done faster than the square of one of them. [REF?]

At a higher level, only the parts of functions with absolute value (norm of the node) above a defined threshold can be retrieved before even processing them; this is used in the scheme presented in the previous section.

Finally at the coarsest level, one can start to perform a SCF calculation with low precision, and use the final result as starting guess for a calculation with higher precision. In contrast to finite basis sets (which usually are not subsets of each others), a multiwavelet function represented with a given precision, can directly be used as a valid representation at a different precision.

@@

MRChem is parallelized using Message Passing Interface (MPI \cite{mpi}) and OpenMP \cite{omp}.

@@

The low level operations (on one or two orbitals or MW trees, and on nodes) are included  in the MRCPP library. The MRCPP will perform tree operations as a set of operations on nodes. Those operations will then be performed in parallel by OpenMP threads. The thread parallelity is transparent for the library user; only the number of threads needs to be set.

@@

In the multiwavelet approach, orbitals are functions represented in a tree structure. The root nodes cover the entire space of the system, and each subdivision defines branches. In order to facilitate the transfer of those functions, the tree structure and the node data are stored in piecewise contiguous memory. This allows the data to be sent efficiently among MPI processes: only the position in memory of the start and end of the pieces are required, and some metadata information. A tree can then be transfered without intermediate buffering and with little overhead. 
The MRCPP library contains functionalities to easily and efficiently transfer MW trees (or orbitals) between MPI processes.

@@

\subsubsection{Task manager} 
In a localized orbital approach, many of the elements of the matrix $\braket{\phi_i|{\hat O} \phi_j}$ will be small. Since the method is inherently adaptive using the MRCPP library, the small elements will be faster to compute. In order to fully exploit this feature, a dynamical distribution of tasks is required.  
In practice the matrix $\braket{\phi_i|{\hat O} \phi_j}$ is partitioned into blocks. Each block represent one task to be performed by a "worker". In a local representation, each node is simply assigned to a separate task. When a "worker" is done with one task it can ask for a new task from the "task manager" until all the tasks are completed.

One single MPI process is reserved for managing the task queue. The only purpose of this process is to keep account of the tasks that have been assigned and the tasks left, and distribute unsolved tasks to the workers. It does only send and receive small amounts of data at a time (a few bytes) and will therefore in practice not be saturated with requests.

@@

\begin{equation}
\label{equ:ex}
  K_{Ex}^k(r) \phi_j = \phi_k(r) \int \frac{\phi^*_k(r')\phi_j(r')}{|r-r'|} \, dr'
\end{equation}

@@

The evaluation of equation (\ref{equ:ex}) can be divided into 3 operations:
\begin{itemize}
\item Product of $\phi^*_k(r')$ and $\phi_j(r')$
\item Application of Poisson operator, $\int \frac{(\phi^*_k(r')\phi_j(r'))}{|r-r'|} \, dr'$
\item Product with $\phi_k(r)$, $\phi_k(r) (\int \frac{\phi^*_k(r')\phi_j(r')}{|r-r'|} \, dr')$
\end{itemize}

@@

The scheme for orthogonalization of orbitals uses the diagonalization of a hermitian matrix and subsequent matrix multiplications which both scale as $0(N_{occ}^3)$
